Pillar 1 - Sequential Pattern Mining

### 2019-02-14 Pillar 1 - Sequential Pattern Mining.Rmd
comments until make_sequences function

# Setup
We load the environment `data_pillar1` which we saved at the end of the data preparation. It contains the data sets `d_course` and `d_transcript`.
```{r loading data}
load("Output/data_pillar_1.RDATA")
```

```{r}
pass_grade <- 5.5
high_grade <- 6.5
```

@@

# Creating Transactions and Sequences
For a first exploration of arules, we conceptualise our framework like this:
transaction = student
item = course

First we transform our data into transaction data. For this, we first create a vector of mandatory courses that we exclude from transcripts. 
## Data Preparation
### Take, fail, low grade
```{r AR data prep 1}
d_transactions <- list()

@@

### Take / not take, pass / fail / not take
```{r transactions not taken}
d_transactions$TPF <- expand.grid(

@@

### Grade
```{r transactions grade}

@@

### Transcript with preceding courses (for SR by hand)
```{r d_transcript_cum}
#
# List of all courses
course_all <- unique(d_transactions$taken_PF_HL$item)

@@

  # item_take
  mutate(
    item_take = paste(item, "take", sep = "_")
  ) %>%

  # course current
  group_by(
    sequenceID,
    eventID
    ) %>%
  summarize_at(
    c("item", "item_take", "item_PF", "item_HL"),
    function(x) list(unique(x))
    ) %>%
  ungroup %>%

  # course current all
  rowwise %>%
  mutate(
    course_current = list(c(item, item_take, item_PF, item_HL))
    ) %>%
  ungroup %>%

  # course past
  group_by(
    sequenceID
    ) %>%
  mutate(
    course_past = Reduce(
      f = c,
      x = course_current,
      accumulate = TRUE
      ) %>%
      lag(
        default = as.character(NA)
        )
    ) %>%
  ungroup %>%

  # course so far, course not so far
  rowwise %>%
  mutate(
    course_past_current = list(union  (course_past, course_current      )),
    course_not_yet      = list(setdiff(course_all , course_past_current )),
    course_not_yet      = list(paste(course_not_yet, "not", sep = "_"   )),
    course_so_far       = list(union  (course_past, course_not_yet))
    ) %>%
  ungroup %>%

  # lhs: past, rhs: present
  select(
    sequenceID,
    lhs = course_so_far,
    rhs = course_current
  )

@@

## Transactions
```{r making transactions, message = FALSE}

@@

## Sequences
```{r making sequences}

@@

#### not take -> fail
```{r SR TPF, eval = FALSE}
#
# SR

x <- d_transcript_cum %>%

  # rhs
  unnest(
    rhs,
    .drop = FALSE
    ) %>%
  filter(
    str_detect(rhs, "low")
    ) %>%

  # lhs
  unnest(
    lhs,
    .drop = FALSE
    ) %>%
  filter(
    str_detect(lhs, "not")
    ) %>%

  # support
  count(
    lhs,
    rhs
    )
```

>>>>

# Setup
We load the environment `data_pillar1` which we saved at the end of the data preparation. It contains the data sets `d_course` and `d_transcript`.
```{r loading data}
load("Output/data_pillar_1.RDATA")
```
We set the thresholds for passing and obtaining a high grade.
```{r}
pass_grade <- 5.5
high_grade <- 6.5
```

@@

# Creating Transactions and Sequences
For a first exploration of arules, we conceptualise our framework like this:
sequence = student
item     = course

## Data Preparation
First we transform our data into transaction data. For this, we first create a vector of mandatory courses that we exclude from transcripts. Furthermore, for courses that span more than one period, we consider only the starting point.

### Take, fail, low grade
For our first transactions we add variables that code whether a course has been passed/failed, or passed with a high grade/low grade. We assign the `Student ID` to the variable `sequenceID`, and the timing of a course (Year,period) to the `eventID`. We identify three different sets of items (course, course_pass/course_fail, course_high/course_low) for the creation of different rules further on. 
```{r AR data prep 1}
d_transactions <- list()

@@

### Take / not take, pass / fail / not take
For our second set of transactions, we come up with all possible combinations of courses and students, then categorise each as, not taken, passed or failed. Items are the courses with their pass/fail/not taken categories (course_pass/course_fail/course_not taken).
```{r transactions not taken}
d_transactions$TPF <- expand.grid(

@@

### Grade
For our last set of transactions we would like to have cumulative probabilities so we can interpret a prediction of a grade 8 as the probability of getting an 8 or lower. 
For this, we add additional rows to the transcripts, corresponding to all the grades above the grade that a student got for a course (e.g. if student_A got a grade_8 for course_B, we would add two rows: student_A/course_B/grade_9 and student_A/course_B/grade_10). Items in these transactions correspond to course_grade.
```{r transactions grade}

@@

### Transcript with preceding courses (for SR by hand)
For our analysis in sequential rule mining, we extract the history of each student, and create a transcript per student, per period, which contains information on the current courses, the courses taken before the current courses, courses taken so far (including current courses) and the courses not taken so far. 
```{r d_transcript_cum}

#
# List of all courses
course_all <- unique(d_transactions$taken_PF_HL$item)

@@

  group_by(
    sequenceID,
    eventID
    ) %>%
  summarize(
    course_current = list(
      unique(item_PF)
      )
    ) %>%

  group_by(
    sequenceID
    ) %>%
  mutate(
    course_past = Reduce(
      f = c, 
      x = course_current, 
      accumulate = TRUE
      ) %>%
      lag(
        default = as.character(NA)
        )
    ) %>%

  rowwise %>% # equivalent to "group by row"
  mutate(
    course_so_far     = list(union  (course_past, course_current)),
    course_not_so_far = list(setdiff(course_all , course_so_far ))
    ) %>%
  ungroup

@@

## Transactions
The function make_transactions takes a data.frame and the item label and converts it into an object of class transaction. 
We create transactions with the different types of data we prepared above (i.e. taken, pass/fail, high/low, NotTaken/pass/fail, Grades).
```{r making transactions, message = FALSE}

@@

## Sequences
The function make_sequence takes a data.frame and the item label and converts it into an object of class sequences (used for extracting sequential rules).
We create sequences with the different types of data we prepared above (i.e. taken, pass/fail, high/low, Grades).
```{r making sequences}

@@

#### not take -> fail
```{r SR TPF, eval = FALSE}

#
# SR by hand

# rhs = current course
d_transcript_cum_current <- d_transcript_cum %>%
  unnest(
    course_current,
    .drop = FALSE
    ) %>%
  rename(
    rhs = course_current
    )

# lhs = course not so far
d_transcript_cum_current %>%
  unnest(
    course_not_so_far,
    .drop = FALSE
    ) %>%
  rename(
    lhs = course_not_so_far
    ) %>%
  count(
    lhs, 
    rhs,
    sort = TRUE
    ) %>%
  filter(str_detect(rhs, "fail"))
```
